---
title: "Interact with Databases Using DBI"
weight: 5
---


The [DBI](http://rstats-db.github.io/DBI/) package provides an interface between R and databases.  Because DBI separates the operations that run against the database from the functions that are exposed to the user and other applications, the same R commands can be used to perform the same operations over a variety of databases.

## When to Use

You should use [dplyr](/dplyr/) for day-to-day data preparation and tidying. The [DBI](http://rstats-db.github.io/DBI/) package is a great tool to help with the rest of database related tasks such as:

- Connecting to and disconnecting from a database
- Retrieving the list of tables and fields
- Adding or removing tables 
- Writing data back to a table
- Running recurring queries 
- Executing stored procedures

The [DBI: R Database Interface](https://cran.r-project.org/web/packages/DBI/vignettes/spec.html) provides a comprehensive review of how DBI works. 


```{r, include = FALSE, cache = TRUE}
library(odbc)
library(config)
library(DBI)

db <- config::get("mssql")

con <- DBI::dbConnect(
  odbc::odbc(),
  Driver = db$Driver,
  Server = db$Server,
  Host = db$Host,
  SVC = db$SVC,
  Database = db$Database,
  Schema = db$Schema,
  UID  = db$UID,
  PWD = db$PWD,
  Port = db$Port)

```

## Connect to a Database

There are two ways to connect to a database.  The first way is to use a **community supported R package**.  Here are the most commonly used packages: 

- [RMySQL]() - Connects to MySQL and MariaDB
- [RPostgreSQL]() - Connects to Postgres and Redshift
- [RSQLite]() - Embeds a SQLite database
- [bigrquery]() - Connects to Google’s BigQuery
- [sparklyr]() - Connect to a Spark context

The second way to connect is to use the [odbc]() package along with an [ODBC]() driver.  Most commercialy available databases provide their own ODBC driver.  

## Open a connection

Use the `dbConnect()` command to open the connection.  Is in this command where we specify the driver, user, password, database name, and other arguments needed to properly start a session with the database.  Conversely, `dbDisconnect()` will close the connection.

A typical `dbConnect()` will look like this:

```{r, eval = FALSE}

library(DBI)
library(odbc)

con <- dbConnect(odbc,
                   Driver    = "SQL Server", 
                   Server    = "localhost",
                   Database  = "airontime",
                   UID       = [My User ID],
                   PWD       = [My Password],
                   Port      = 1433)

```

## Tables & Fields

If you are using the RStudio IDE, then exploring an open database connection is as easy as expanding and drilling down into the schema using the Connections tab. 

Using code, to get a list of the tables in an existing database connection, use `dbListTables()`

```{r, eval = FALSE}
dbListTables(con)
```

```{r, echo = FALSE, cache = TRUE}
dbListTables(con)[1:6]
```

For a list of the fields in a particular table inside the database connection use `dbListFields()`

```{r, echo = TRUE, cache = TRUE}
dbListFields(con, "flights")
```

## Adding and Removing Tables

The `dbWriteTable()` and `dbRemoveTable()` can be used to perform their respective operations.  The user will need to have the proper database rights to be able to run these operations.

A data frame can be passed to `dbWriteTable()` to create and simultaneously populate the table:

```{r, eval = FALSE , cache = TRUE}
library(nycflights13)
dbWriteTable(con, "flights", flights)
```


## Running a Query

The `dbGetQuery()` command allows us to write queries and retrieve the results. How to best use this command is examined in this article: [Run Queries Safely](/run-queries-safely/)




