---
title: RStudio Drivers
weight: 5
---

## Install RStudio's ODBC drivers 

The [odbc-install](https://github.com/rstudio/odbc-install) repository has a script that simplifies the installation of the ODBC drivers in the client machine.

### Red Hat Enterprise Linux/CentOS Linux (6.0+)

There are two system dependencies that need to be installed prior to installing the drivers

```{}
sudo yum install wget
sudo yum install unixODBC
``` 

The following script will download and install the drivers:

```{}
# Placeholder path
wget -O - https://s3-us-west-2.amazonaws.com/rstudio-odbc/...
```


## Connecting to a Database using the odbc package

The [odbc](https://github.com/rstats-db/odbc) package provides a DBI-compliant interface to ODBC drivers.  Drivers to a wide assortment of database types are provided as part of RStudio's commercial products.  

**Why is a DBI compliant interface important? ** Apart from enabling the analyst to leverage [DBI](http://rstats-db.github.io/DBI/) for their analysis, it also opens the door for a [dplyr](https://github.com/tidyverse/dplyr) back end to be developed which makes data wrangling transparent to the analysts since they can continue using the same *dplyr* commands.

```{r, eval = FALSE}
devtools::install_github("rstats-db/odbc")
```

Use **dbConnect** to access the database. Passing the values from the **config** file enables us to protect the credentials and abstracts the server and database name in case we need to promote the code to point from a test to a production database.

```{r, message=FALSE}
library(odbc)
library(config)
library(DBI)

con <- dbConnect(odbc::odbc(), 
  .connection_string = paste0("Driver={sqlserver};
                              Server=", config::get("datawarehouse")$server ,";
                              Port=1433;
                              Database=", config::get("datawarehouse")$database ,";
                              UID=", config::get("datawarehouse")$uid ,";
                              PWD=", config::get("datawarehouse")$pwd ,";"))

```
