---
title: Securing Credentials using the keyring package
weight: 7
---

The [keyring](https://github.com/gaborcsardi/keyring) package enables us to access the operating system's credential store. 

```{r, eval = FALSE}
devtools::install_github("gaborcsardi/keyring")
```

In this example, we will create a new key ring called *rstudio.odbc* and add a single key with the server path and credentials. Additional servers can be added to the same key ring by adding additional keys.  

The way **keyring** retrieves data is by passing the Key Ring name and a **service** name.  In this method, we will use the server path as the **service** so it will act as our key to retrieve user name and password to use for that particular server. 

To prevent passing the server path in our code, we will use another package called **config**, which is discussed in the next section.

```{r, eval = FALSE}
keyring::keyring_create("rstudio.odbc")

keyring::key_set_with_value(service = "sol-eng-sqlserv.cihykudhzbgw.us-west-2.rds.amazonaws.com", 
                            username = "rstudioadmin", 
                            password = "ABCd4321", 
                            keyring = "rstudio.odbc")
```