---
title: Secure Credentials
weight: 7
---

The ability to share code with others is one of the most appealing features of R. It is very important to no


When a database connection is part of our code, it it very important to avoid leaving as plain text the credentials we used. Ideally, credentials should be encrypted. 

In this article we will cover five ways to make sure that your credentials are safe.  Starting with the best method at the top, here are the methods that we will cover:


- Integrated credentials



- Encrypt credentials

- Use a configuration file

- Prompt for Credentials

- Use Environment variables



## Integrated credentials

## Encrypt credentials

The [keyring](https://github.com/gaborcsardi/keyring) package enables us to secure any credential by leveraging the operating system's credential store. We will use this capability to securely store database credentials.

```{r, eval = FALSE}
devtools::install_github("gaborcsardi/keyring")
```

### Keyrings and keys

Inside the credential store there are `keyrings` that contain multiple `keys`.  A `keyring` is secured with a single password, which then grants acccess to to all of its `keys`.  In our case, we will store a single database connection credential per `key`. We will use the default keyring, which is automatically un-locked when a user signs in.

### Our first key

A `key` has four main attributes:

- `service` - This is the key's unique identifier, we will use the database server's path for our keys.
- `keyring` - The key's 'parent' keyring. If not specified, the default keyring is used.
- `username`
- `password` 

The way `keyring` retrieves data is by passing the `keyring` name and a `service` name. If no `keyring` is passed, the default keyring is used.

We only create the key one time:

```{r, eval = FALSE}
keyring::key_set_with_value(service = "mssql-database", 
                            username = "newuser", 
                            password = "abc123")
```

### Retrieve credentials

The `key_list()` command is used to retrieve the `username`:

```{r}
keyring::key_list("mssql-database")
```

To extract the only the `username`, so as to pass it inside the connection string, use:

```{r}
keyring::key_list("mssql-database")[1,2]
```

To extract the `password`, use `key_get()`:

```{r}
keyring::key_get("mssql-database")
```

These functions can be used to supply the database credentials without storing them in plain text or an environment variable:

```{r, eval = FALSE}
con <- dbConnect(odbc::odbc(), 
  "SQLServer", # A driver defined in odbcinst.ini
  Server="mssql-database",
  Port=1433,
  Database="default",
  UID= keyring::key_list("mssql-database")[1,2],
  PWD= keyring::key_get("mssql-database"))
```

The default keyring is unlocked anytime the user is signed in. If a new keyring is created and used, the Operating System will prompt the user for the `keyring` password when the code executes.


## Use a configuration file

## Prompt for Credentials

## Use Environment variables

